WASP-4 is Accelerating Towards the Sun

### 2020-02-09 paper/ms.tex
fix column width

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{\today}
\revised{---}
\accepted{---}
\submitjournal{AAS journals.}
\shorttitle{WASP-4 is accelerating towards the Earth}

@@

\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be decreasing at
  a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.
  Proposed explanations for the period change include
  tidal orbital decay, apsidal precession, and acceleration of the system
  along the line of sight.
  To investigate further, we performed new radial velocity measurements and
  speckle imaging of WASP-4.
  The radial-velocity data show that the system is
  accelerating towards the Sun at a rate of
  $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.
  The associated Doppler effect should cause
  the apparent period to shrink at a rate of
  $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.
  Thus, the observed change in the transit period is mostly
  or entirely produced by the line-of-sight acceleration of the system.
  This acceleration is probably caused by a wide-orbiting companion
  of mass 10-200$\,M_{\rm Jup}$ and orbital distance 10-100$\,$AU,
  based on the magnitude of the radial-velocity trend and the non-detection of any
  companion in the speckle image.  We expect that 1 out of 6 hot Jupiters
  will show apparent period changes of the same order of magnitude as WASP-4,
  based on the hot-Jupiter companion statistics of  
  \citet{knutson_friends_2014}. Continued radial velocity
  monitoring of hot Jupiters is therefore essential to distinguish
  the effects of tidal orbital decay or apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

% FIXME : verify with Michael Gillon that this is true. (P. Evans
% done)
Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m telescope at La Silla, Chile,
and the South African Astronomical Observatory 1.0m telescope.
% Five of the twenty-two occurred at simultaneous epochs as transits
% previously reported by \citet{hoyer_tramos_2013}.
Additional timing measurements were also reported recently by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations.
We included twelve of their ``high quality'' transit times from
the TRAPPIST telescope (six transits), El~Sauce (four transits), and
\citet{petrucci_no_2013}.  For TRAPPIST and El~Sauce, we verified with
the original observers that correct barycentric and leap-second
corrections had been performed (M.~Gillon, P.~Evans, priv{.} comm{.}).
We omitted the fourteen remaining \citeauthor{baluev_2019}
ETD\footnote{\url{http://var2.astro.cz/ETD}} times due to ambiguity in
whether leap-second corrections had been performed.
We did not include in our analysis the four occultation times tabulated by
\citetalias{bouma_wasp4b_2019}, because of the large timing
uncertainties and negligible statistical power.

@@

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29.  On each night, we acquired three sets of $1000\times 60$\,$msec exposures.
If a companion is present, the autocorrelation
functions of these speckle images would reveal a characteristic interference
pattern. This pattern in then used to determine the properties of the
detected companion and produce a reconstructed image.  Using the
reconstructed speckle images, contrast curves are produced to determine the
5-$\sigma$ detection limits (see \citealt{howell_speckle_2011}).  No
companions were detected. The data from the second night, which had better
seeing (0.6$''$), led to the most constraining limit.  The
832$\,$nm limits were the most useful, given that any faint
companion would likely be redder than the host star.  Therefore,
we opted to use
the 832$\,$nm September 28-29 contrast limits in the analysis described below.

@@

Our initial model for the radial velocity data was a circular orbit
plus instrument offsets, ``jitter'' values (explained below), and a long-term linear trend
\citep[][\texttt{radvel}]{fulton_radvel_2018}.  We set Gaussian priors
on the orbital period and time of inferior conjunction using the
values from Table~4 of \citetalias{bouma_wasp4b_2019}.
We assumed the orbit to be circular based on previous studies
that placed stringent upper limits on the eccentricity \citep{beerer_secondary_2011,knutson_friends_2014,bonomo_gaps_2017}.
The free parameters were the orbital velocity semi-amplitude, the instrument
zero-points, an additive ``white noise'' instrument jitter for each
instrument,
% JNW: Do you mean "added in quadrature to the formal uncertainties"? That's the usual approach.
and a linear ($\dot{v_{\rm r}}$) acceleration term.  (Without the linear
trend, the model is a much poorer fit with $\Delta {\rm BIC}=73$.)

>>>>

%% Reintroduced the \received and \accepted commands from AASTeX v5.2.
%% Add "Submitted to " argument.
\received{\today}
\revised{---}
\accepted{---}
\submitjournal{AAS journals.}
\shorttitle{WASP-4 is accelerating towards the Sun}

@@

%TODO FIXME
\begin{abstract}
  The orbital period of the hot Jupiter WASP-4b appears to be
  decreasing at a rate of $-8.64 \pm 1.26~{\rm msec}\,{\rm yr}^{-1}$,
  based on transit-timing measurements spanning 12 years.  Proposed
  explanations for the period change include tidal orbital decay,
  apsidal precession, and acceleration of the system along the line of
  sight.  To investigate further, we performed new radial velocity
  measurements and speckle imaging of WASP-4.  The radial-velocity
  data show that the system is accelerating towards the Sun at a rate
  of $-0.0422\pm 0.0028\,{\rm m}\,{\rm s}^{-1}\,{\rm day}^{-1}$.  The
  associated Doppler effect should cause the apparent period to shrink
  at a rate of $-5.94 \pm 0.39~{\rm msec}\,{\rm yr}^{-1}$, comparable
  to the observed rate.  Thus, the observed change in the transit
  period is mostly or entirely produced by the line-of-sight
  acceleration of the system.  This acceleration is probably caused by
  a wide-orbiting companion of mass 10-200$\,M_{\rm Jup}$ and orbital
  distance 10-100$\,$AU, based on the magnitude of the radial-velocity
  trend and the non-detection of any companion in the speckle image.
  We expect that 1 out of 6 hot Jupiters will show apparent period
  changes of the same order of magnitude as WASP-4, based on the
  hot-Jupiter companion statistics of  \citet{knutson_friends_2014}.
  Continued radial velocity monitoring of hot Jupiters is therefore
  essential to distinguish the effects of tidal orbital decay or
  apsidal precession from line-of-sight acceleration.
\end{abstract}

@@

% FIXME : update wording
Most of these data are identical to the data presented by
\citetalias{bouma_wasp4b_2019}.  Twenty-two new times reported by
\citet{southworth_transit_2019} are included.  These transits were
observed from the 3.58m New Technology Telescope and the Danish 1.54m
telescope at La Silla, Chile, and the South African Astronomical
Observatory 1.0m telescope.
% Five of the twenty-two occurred at simultaneous epochs as transits
% previously reported by \citet{hoyer_tramos_2013}.
Additional timing measurements were also reported recently by
\citet{baluev_2019}, based on a homogeneous analysis of archival
ground-based observations.  We included twelve of their ``high
quality'' transit times from the TRAPPIST telescope (six transits),
El~Sauce (four transits), and \citet{petrucci_no_2013}.  For TRAPPIST
and El~Sauce, we verified with the original observers that correct
barycentric and leap-second corrections had been performed (M.~Gillon,
P.~Evans, priv{.} comm{.}).  We omitted the fourteen remaining
\citeauthor{baluev_2019} ETD\footnote{\url{http://var2.astro.cz/ETD}}
times due to ambiguity in whether leap-second corrections had been
performed.  We did not include in our analysis the four occultation
times tabulated by \citetalias{bouma_wasp4b_2019}, because of the
large timing uncertainties and negligible statistical power.

@@

We observed WASP-4 twice, on the night of September 11-12 with
relatively poor seeing (1.2$''$) and also on the night of September
28-29.  On each night, we acquired three sets of $1000\times
60$$\,$msec exposures.  If a companion is present, the autocorrelation
functions of these speckle images would reveal a characteristic
interference pattern. This pattern in then used to determine the
properties of the detected companion and produce a reconstructed
image.  Using the reconstructed speckle images, contrast curves are
produced to determine the 5-$\sigma$ detection limits (see
\citealt{howell_speckle_2011}).  No companions were detected. The data
from the second night, which had better seeing (0.6$''$), led to the
most constraining limit.  The 832$\,$nm limits were the most useful,
given that any faint companion would likely be redder than the host
star.  Therefore, we opted to use the 832$\,$nm September 28-29
contrast limits in the analysis described below.

@@

Our initial model for the radial velocity data was a circular orbit
plus instrument offsets, ``jitter'' values (explained below), and a
long-term linear trend \citep[][\texttt{radvel}]{fulton_radvel_2018}.
We set Gaussian priors on the orbital period and time of inferior
conjunction using the values from Table~4 of
\citetalias{bouma_wasp4b_2019}.  We assumed the orbit to be circular
based on previous studies that placed stringent upper limits on the
eccentricity
\citep{beerer_secondary_2011,knutson_friends_2014,bonomo_gaps_2017}.
The free parameters were the orbital velocity semi-amplitude, the
instrument zero-points, an additive ``white noise'' instrument jitter
for each instrument, and a linear ($\dot{v_{\rm r}}$) acceleration
term.  (Without the linear trend, the model is a much poorer fit with
$\Delta {\rm BIC}=73$.)
